Vagueness, Noise, and Signaling

### 2015-05-14 paper/version_03/paper.tex
PHoS paper Derivation RDD added; major edits and shortening relation to other accounts; figure Voronoi-language added

\newcommand{\impairment}{\ensuremath{\alpha}} % impairment
\newcommand{\toler}{\ensuremath{\beta}} % tolerance
\newcommand{\ns}{\ensuremath{n_s}} % number of states

@@

\begin{abstract}
  Signaling games are popular models for studying the evolution of
  meaning, but typical approaches do not incorporate vagueness as a
  feature of successful signaling.  Complementing recent like-minded
  models, we introduce the replicator diffusion dynamic as a special
  case of the replicator mutator dynamic that combines evolutionary
  pressure on optimal signal use with stochastic noise in the
  processing of similar stimuli. Applying this new dynamic to a
  generalization of Lewis' signaling games, we show that stochastic
  imprecision leads to vague, yet by-and-large communicative efficient
  signal use, and, moreover, that it unifies evolutionary outcomes and helps
  avoid suboptimal categorization.
\end{abstract}

@@

A number of authors have recently tried to explain why vagueness
evolved as something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of
limitations in information processing or particular learning
strategies
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This
paper contributes to the latter line of thought. Concretely, we
introduce the replicator diffusion dynamics---a novel variant of the
replicator mutator dynamic---that integrates stochastic noise on the
differential confusability of similar stimuli. The dynamic proposed
here generalizes and complements previous like-minded accounts. We
show that stochastic noise can not only lead to vague, yet relatively
efficient signal use, but also unify evolutionary outcomes and
help avoid suboptimal categorization.

The next section introduces the background against which the work
presented here can be appreciated. Section~\ref{sec:repl-diff-dynam}
introduces the replicator diffusion dynamic, which is then explored in Section~\ref{sec:exploring-rdd}.
Section~\ref{sec:discussion} reflects and compares
our approach to others.
In Appendix~\ref{sec:diffusion-as-special} we give a formal elaboration on the relation between the replicator diffusion dynamic and the replicator mutator dynamic.
This finding is not necessary to understand the main points of this paper regarding vagueness, hence its inclusion in an appendix.
We nevertheless include it since it is related, conceptually interesting, and potentially valuable for future work related to the model introduced here.

@@

A sim-max game consists of a set of states $\States$, a set of
messages $\Messgs$ with much fewer messages than states, a prior
probability distribution $\Pr \in \Delta(\States)$ that gives the
occurrence probabilities of states, a similarity metric on states
$\similarity \mycolon \States \times \States \rightarrow \mathds{R}$,
and a utility function $\Utils \mycolon \States \times \States
\rightarrow \mathds{R}$. We identify the receiver's acts with the
states of the world, so that the game is one of guessing the actual
state, so to speak. We also assume that sender's and receiver's
interests are alike, so we only have one utility function. We do not
consider message costs, so utilities only depend on the actual state
and the receiver's response. The utility function should be a
monotonically decreasing function of similarity.

@@

\begin{figure}
  \centering
  \missingfigure{Please include here something like Figure 4 in your paper with Elliot.}
  \caption{Example of Voronoi language for the unit interval.}
  \label{fig:Voronoi-language-example}
\end{figure}

@@

The formulation of the \rdd given and explored here is in terms of
behavioral strategies, not in terms of mixed strategies. Dynamics on
behavioral strategies assume that agents can adapt
their behavior locally, so to speak, i.e., independently at each
choice point. This greatly reduces the complexity of the dynamic and
simplifies numerical simulations. But it also points towards an
interpretation of the process as cultural evolution, rather than
biological evolution. Signaling strategies are not inherited as a
whole, but adapted piece-meal, e.g., by conditional imitation of
``local'' behavior at a given choice point (more below). Still, the
approach we suggest here is compatible with a formulation in terms of
mixed strategies and an interpretation as biological evolution as
well. In Appendix~\ref{sec:diffusion-as-special}, we
translate our initial formulation of the \rdd
into one using mixed strategies. Doing so identifies the \rdd as a
special case of the \rmd.

@@

Fix a signaling game with finite states $\States$, messages $\Messgs$
and acts $\Acts$. Let $\Pr(\cdot) \in \Delta{\States}$ be the prior
distribution over states and $\Util_{\sen,\rec} \mycolon \States
\times \Messgs \times \Acts \rightarrow \mathds{R}$ the sender's and
receiver's utility functions. A behavioral strategy is a function that
maps an agent's choice points to a probability distribution over
available choices. The sender's behavioral strategies are functions
$\Sstrat \in \Delta(\Messgs)^\States$, thus mapping each state $\state \in \States$ to a probability of each message $\messg \in \Messgs$ being sent; the receiver's are functions
$\Rstrat \in \Delta(\Acts)^\Messgs$, thus mapping each message $\messg \in \Messgs$ to a probability of each state $\state \in \States$ being interpreted by the receiver given $\messg$. The expected utility of choices
at each choice point is:
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act, \messg, \Sstrat) & = \sum_{\state \in
    \States} \Pr(\state) \cdot \Sstrat(\messg \probbar \state) \cdot
  U_\rec(\state, \messg, \act) \,.
\end{align*}
This means that the expected utility for the sender of choosing message $\messg \in \Messgs$, given a state $\state \in \States$ and a receiver strategy $\Rstrat$ is the sum, for each act $\act \in \Acts$, of the probability the receiver will choose act $\act$ given message $\messg$ times the actual utility of using act $\act$ for state $\state$.
Conversely, the expected utility for the receiver of choosing act $\act \in \Acts$, given a message $\messg \in \Messgs$ and a sender strategy $\Sstrat$ is the sum, for each state $\state \in \States$, of the prior probability of state $\state$ occurring, times the probability the sender will choose message $\messg$ given state $\state$, times the actual utility of using act $\act$ for state $\state$.

The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg, \state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act, \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$
in such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state) & = \frac{\Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)} {\Phi(\state,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\act \probbar \messg) & = \frac{\Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}
The \rd in behavioral strategies describes the most likely path of
development of average behavior in a population
of agents when behavioral adaptations occur independently at each
choice point, e.g., when agents adopt the behavior of another agent
\emph{at a given choice point} if that ``local'' behavior is
``locally'' better than their own (see
\citeauthor{Sandholm2013:Population-Game}
\citeyear{Sandholm2013:Population-Game} for agent-level
interpretations of the replicator dynamics and
\citeauthor{Cressman2003:Evolutionary-Dy}
\citeyear{Cressman2003:Evolutionary-Dy} for applications to extensive
form games and behavioral strategies).

@@

The discrete-time replicator diffusion dynamic takes the replicator
dynamic as basic, but factors in the confusion of states at each step
in a sequential manner:
\begin{align*}
  \RDD(\Sstrat) & = \Diff_C(\RD(\Sstrat)) &   \RDD(\Rstrat) & = \Diff_C(\RD(\Rstrat)) \,.
\end{align*}
This is equivalent to the following, perhaps more transparent
formulation:\footnote{To see this, start with $\RDD(\Sstrat)(\messg_k
  \probbar \state_i)$. If $\RDD(\Sstrat)$ is considered as a matrix,
  we can rewrite this as $(RDD(\Sstrat)_{ik}$. By
  Equation~(\ref{eq:confusion-function}), we get $(C \
  \RD(\Sstrat))_{ik}$, which by definition of matrix multiplication is
  $\sum_j C_{ij} \cdot \RD(\Sstrat)_{jk}$. Substituting the definition
  of $\RD(\Sstrat)$ completes the derivation. Similarly for the
  receiver.}
\begin{align*}
  \RDD(\Sstrat)(\messg_k \probbar \state_i) & = \sum_{j} C_{ij} \cdot
  \frac{\Sstrat(\messg_k \probbar \state_j) \cdot
    \EU(\messg_k, \state_j,\Rstrat)}
  {\Phi(\state_j,\Sstrat, \Rstrat)} \\
    \RDD(\Rstrat)(\state_i \probbar \messg_k) & = \sum_{j} C_{ji} \cdot
  \frac{\Rstrat(\state_j \probbar \messg_k) \cdot
    \EU(\state_j, \messg_k,\Sstrat)} {\Phi(\messg_k,\Sstrat, \Rstrat)}  \,.
\end{align*}

The most straightforward interpretation of this sequential definition
is that the \rdd captures the most likely path of development of
aggregate behavior in a society, in which behavioral changes occur
independently at each choice point, as described by the \rd for
behavioral strategies. Whatever strategy an agent does actually hold,
its realization is perturbed by imperfect discriminability of states
in such a way that the expected overall aggregate behavior is the
diffusion of the strategies that the agents would play in a noise-free
world. Any agent-level interpretation of the \rd in behavioral
strategies seems \emph{prima facie} compatible with this
interpretation.

The sequential definition highlights that other (discrete-time)
evolutionary dynamics can be subjected to state-confusability in an
analogous fashion. It is easy, for instance, to perturb a
best-response dynamic in case we wanted to model cognizing, rational
agents with a noisy perceptual system. Our choice of the \rd as a
carrier dynamic is motivated by its generality and austerity, being
compatible with a variety of implicit agent-level behavior, without
presupposing extravagant higher cognitive abilities
\citep[c.f.][]{Sandholm2013:Population-Game}. Moreover, by using the
\rd as a carrier for diffusion, we actually obtain a special case of a
behavioral-strategy version of the replicator mutator dynamics. This
is important to know because it enables similar applications of
diffusion from perceptual confusion also in models of biological
evolution.
We worked out the relation between the \rdd and the \rmd, but given that it is rather technical and not strictly necessary to appreciate the main point of the paper, we defer it to Appendix~\ref{sec:diffusion-as-special}.
In the following section we explore the long-term effects of using the \rdd.

@@

The previous section argued that the \rdd can be conceptualized as a
two-step procedure: we first calculate where the replicator dynamic
would take us, then apply diffusion. To understand the \rdd better, we
isolate the novel diffusion part in
Section~\ref{sec:iterated-diffusion}, before exploring how diffusion
and fitness-based selection interact in Section~\ref{sec:simulations}.

@@

Confusion of states should intuitively be a function of their
perceptual similarity: the more similar two states are, the more likely they could be mistaken for each other.
To make this concrete, let us assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. The
distance $\card{\mystate{i} - \mystate{j}}$ is the objective, physical
similarity between two states $\mystate{i}$ and
$\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \impairment) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \impairment^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\impairment \ge 0$ is an impairment, imprecision or indiscriminability
parameter\footnote{We use these terms interchangeably throughout the paper.}.
When $\impairment=0$ agents perfectly discriminate between
states; when $\impairment \rightarrow \infty$ agents cannot
discriminate states at all. Figure~\ref{fig:NosofskySim} gives an
impression of Nosofsky-similarity for different parameter
values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different
areas of the state space, but we stick with Nosofsky's similarity
function for the time being, because it is mathematically simple, yet
an established notion in mathematical psychology.

@@

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\impairment$. For $\impairment = 0$, the confusion matrix has $C_{ii}
= 1$ and $C_{ij} = 0$ for $i \neq j$. For $\impairment > 0$, $C$ is positive, i.e., $C_{ij} >0$ for
all $i$ and $j$. For $\impairment \rightarrow \infty$, we find $C_{ij}
\rightarrow \nicefrac{1}{\card{\States}}$ for all $i$ and $j$.

If states are confused on average with a probability proportional to
their similarity, repeated application of the operation in
Equation~(\ref{eq:confusion-function}) causes behavioral dispositions
gradually to diffuse along a gradient of similarity of
states. Consequently, iterated diffusion leads to a smoothing out and
an eventual equalization of behavioral strategies, for both the sender
and the receiver. After $n$ steps of iterated diffusion an initial
sender strategy $\Sstrat$ will be $C^n \Sstrat$, and an initial
receiver strategy $\Rstrat$ will be $\Rstrat C^n$. If we assume that
$\impairment > 0$, $C$ is a positive row-stochastic matrix (each state
could in principle be confused as any other state, albeit with
possibly a very low probability). It then follows from the
Perron-Frobenius theorem that the limit $C^\infty = \lim_{n
  \rightarrow \infty} C^n$ exists (it is the Perron-projection) and
all of its rows are identical. That is why, in the limit of iterated
diffusion, $C^\infty \Sstrat$ has identical rows (messages are sent
with the same probability in each state), and $\Rstrat C^\infty$
likewise has identical rows (every message is interpreted
equally). All of this is in line with the intuition that diffusion of
behavior under confusion of states iteratively equalizes behavior for
similar states; if all states can be confused for one another in
principle, behavior smoothes out entirely in the limit.

@@

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\impairment = 0$, the
\rdd reduces to the \rd. If $\impairment \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion process that we looked at in the previous
section. We will show presently that for reasonable in-between levels
of imprecision $\impairment$, the \rdd can lead to reasonably
efficient, yet vague signaling behavior. Suitable levels of
imprecision can have further accelerating and, surprisingly, unifying
effects on meaning evolution.

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind3098.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0$}
    \label{fig:example_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind3452.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0.1$}
    \label{fig:example_stratsB}
  \end{subfigure}

>>>>

\newcommand{\imprecision}{\ensuremath{\alpha}} % imprecision
\newcommand{\toler}{\ensuremath{\beta}} % tolerance
\newcommand{\ns}{\ensuremath{n_s}} % number of states

@@

\begin{abstract}
  Signaling games are popular models for studying the evolution of meaning, but typical
  approaches do not incorporate vagueness as a feature of successful signaling.  Complementing
  recent like-minded models, we introduce the replicator diffusion dynamic that combines
  evolutionary pressure on optimal signal use with stochastic noise in the processing of
  similar stimuli. Applying this new dynamic to a generalization of Lewis' signaling games, we
  show that stochastic imprecision leads to vague, yet by-and-large efficient signal use, and,
  moreover, that it unifies evolutionary outcomes and helps avoid suboptimal categorization.
\end{abstract}

@@

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing or particular learning strategies
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce the replicator diffusion
dynamics that integrates stochastic noise on the differential confusability of similar
stimuli. The dynamic proposed here generalizes and complements previous like-minded
accounts. We show that stochastic noise can not only lead to vague, yet relatively efficient
signal use, but can also unify evolutionary outcomes and help avoid suboptimal categorization.

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces the replicator diffusion dynamic,
which is then explored in Section~\ref{sec:exploring-rdd}.  Section~\ref{sec:discussion}
reflects and compares our approach to others. Two appendices cover technical material that is
not strictly necessary to appreciate the main conceptual points of the paper.
Appendix~\ref{sec:repl-mutat-dynam} shows that out replicator diffusion dynamic is a special
case of the well-known replicator mutator dynamics. Appendix~\ref{sec:deriv-rdd-imit} gives a
derivation of our dynamic as the most-likely path of a population whose agents imitate
successful behavior without being aware of perceptual noise themselves.

@@

A sim-max game consists of a set of states $\States$, a set of messages $\Messgs$ with much
fewer messages than states, a prior probability distribution $\Pr \in \Delta(\States)$ such
that $\Pr(\state)$ gives the probability that state $\state$ occurs, a similarity
metric\footnote{A metric is a function of distance between any two points in a given space. It
  should satisfy certain axioms that one would intuitively expect from the words ``similarity''
  and ``distance'' alone, but these details do not matter for the purposes of this paper.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function $\Utils \mycolon \States \times \States \rightarrow
\mathds{R}$ such that $\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver
for a play with actual state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We
identify the receiver's acts with the states of the world, so that the game is one of guessing
the actual state, so to speak. The utility function should be a monotonically decreasing
function of similarity.

@@

  \begin{tikzpicture}[scale=3]

          \draw[-,thick] (0,0) -- (1.5,0);

          \draw (0,0) node {\color{red}{\big{[}}};

          \draw (.75,0) node {\color{red}{\big{)}}};

          \draw (.76,0) node {\color{red}{\big{(}}};

          \draw (1.5,0) node {\color{red}{\big{]}}};

            \draw (.325,.15) node
            {\color{red}{$\Spure(\mymessg{1}  \probbar \cdot)$}};

            \draw (1.075,.15) node
            {\color{red}{$\Spure(\mymessg{2} \probbar \cdot)$}};

            \node at (.375,0) (node1) {\color{blue}{$\bullet$}};

            \node at (.375,-.3) (label1)
            {\color{blue}{${\Rpure}(\cdot \probbar \mymessg{1})$}};

            \draw[-,color=blue] (node1) -- (label1);

            \node at (1.125,0) (node2) {\color{blue}{$\bullet$}};

            \node at (1.125,-.3) (label2)
            {\color{blue}{${\Rpure}(\cdot \probbar \mymessg{2})$}};

            \draw[-] (node2) -- (label2);

            \draw (0,-.2) node {$0$};

            \draw (.75,-.2) node {$0.5$};

            \draw (1.5,-.2) node {$1$};

          \end{scope}

        \caption{Example of a Voronoi language on $\States = [0,1]$. The (pure) sender strategy
          $\Spure \in \Messgs^\States$ uses one signal for the lower half, and another for the
          upper half of the unit interval. The (pure) receiver strategy $\Rpure \in
          \States^\Messgs$ selects the central elements in the respective intervals. See
          \citet{JagerMetzger2011:Voronoi-Languag} for further details.}
  \label{fig:Voronoi-language-example}
\end{figure}

@@

The formulation of the \rdd given and explored here is in terms of behavioral strategies, not
in terms of mixed strategies. Dynamics on behavioral strategies assume that agents can adapt
their behavior locally, so to speak, i.e., independently at each choice
point. Section~\ref{sec:repl-mutat-dynam} elaborates on the
difference. Section~\ref{sec:relating-dynamics} shows how to translate our initial formulation
of the \rdd into one using mixed strategies. Doing so identifies the \rdd as a special case of
the \rmd.

Our focus on behavioral strategies greatly reduces the complexity of the dynamic and simplifies
numerical simulations. But it also points towards an interpretation of the process as cultural
evolution, rather than biological evolution. Signaling strategies are not inherited as a whole,
but adapted piece-meal, e.g., by conditional imitation of ``local'' behavior at a given choice
point. Appendix~\ref{sec:deriv-rdd-imit} gives one possible derivation of our dynamic as the
most likely path in a population of agents that imitate successful behavior while being
entirely unaware of perceptual imprecision themselves. Still, the possibility to formulate the
\rdd as a special case of the \rmd (Appendix~\ref{sec:diffusion-as-special}) also suggests that
an interpretation as a process of biological evolution may be feasible.

@@

Fix a signaling game with finite states $\States$, messages $\Messgs$
and acts $\Acts$. Let $\Pr(\cdot) \in \Delta{\States}$ be the prior
distribution over states and $\Util_{\sen,\rec} \mycolon \States
\times \Messgs \times \Acts \rightarrow \mathds{R}$ the sender's and
receiver's utility functions. A behavioral strategy is a function that
maps an agent's choice points to a probability distribution over
available choices. The sender's behavioral strategies are functions
$\Sstrat \in \Delta(\Messgs)^\States$, thus mapping each state $\state \in \States$ to a probability of each message $\messg \in \Messgs$ being sent; the receiver's are functions
$\Rstrat \in \Delta(\Acts)^\Messgs$, thus mapping each message $\messg \in \Messgs$ to a probability of each state $\state \in \States$ being interpreted by the receiver given $\messg$. The expected utility of choices
at each choice point is:
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act, \messg, \Sstrat) & = \sum_{\state \in
    \States} \Pr(\state) \cdot \Sstrat(\messg \probbar \state) \cdot
  U_\rec(\state, \messg, \act) \,.
\end{align*}
This means that the expected utility for the sender of choosing message $\messg \in \Messgs$,
given tate $\state \in \States$ and a receiver strategy $\Rstrat$ is the sum, for each act
$\act \in \Acts$, of the probability that the receiver will choose act $\act$ given message
$\messg$ times the actual utility of using act $\act$ for state $\state$.  The expected utility
for the receiver of choosing act $\act \in \Acts$, given a message $\messg \in \Messgs$ and a
sender strategy $\Sstrat$ is the sum, for each state $\state \in \States$, of the prior
probability of state $\state$ occurring, times the probability the sender will choose message
$\messg$ given state $\state$, times the actual utility of using act $\act$ for state $\state$.

The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg, \state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act, \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$
in such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state) & = \frac{\Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)} {\Phi(\state,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\act \probbar \messg) & = \frac{\Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}
The \rd in behavioral strategies describes the most likely path of development of average
behavior in a population of agents when behavioral adaptations occur independently at each
choice point, e.g., when agents adopt the behavior of another agent \emph{at a given choice
  point} if that ``local'' behavior is ``locally'' better than their own (see
\cite{Sandholm2010:Population-Game} for agent-level interpretations of the replicator dynamics
and \citeauthor{Cressman2003:Evolutionary-Dy} \citeyear{Cressman2003:Evolutionary-Dy} for
applications to extensive form games and behavioral strategies).

@@

The discrete-time replicator diffusion dynamic takes the replicator
dynamic as basic, but factors in the confusion of states at each step
in a sequential manner:
\begin{align}
  \label{eq:RDD-Def}
  \RDD(\Sstrat) & = \Diff_C(\RD(\Sstrat)) &   \RDD(\Rstrat) & = \Diff_C(\RD(\Rstrat)) \,.
\end{align}
This is equivalent to the following, perhaps more transparent
formulation:\footnote{To see this, start with $\RDD(\Sstrat)(\messg_k
  \probbar \state_i)$. If $\RDD(\Sstrat)$ is considered as a matrix,
  we can rewrite this as $(RDD(\Sstrat)_{ik}$. By
  Equation~(\ref{eq:confusion-function}), we get $(C \
  \RD(\Sstrat))_{ik}$, which by definition of matrix multiplication is
  $\sum_j C_{ij} \cdot \RD(\Sstrat)_{jk}$. Substituting the definition
  of $\RD(\Sstrat)$ completes the derivation. Similarly for the
  receiver.}
\begin{align*}
  \RDD(\Sstrat)(\messg_k \probbar \state_i) & = \sum_{j} C_{ij} \cdot
  \frac{\Sstrat(\messg_k \probbar \state_j) \cdot
    \EU(\messg_k, \state_j,\Rstrat)}
  {\Phi(\state_j,\Sstrat, \Rstrat)} \\
    \RDD(\Rstrat)(\state_i \probbar \messg_k) & = \sum_{j} C_{ji} \cdot
  \frac{\Rstrat(\state_j \probbar \messg_k) \cdot
    \EU(\state_j, \messg_k,\Sstrat)} {\Phi(\messg_k,\Sstrat, \Rstrat)}  \,.
\end{align*}

The most straightforward interpretation of this sequential definition is that the \rdd captures
the most likely path of development of aggregate behavior in a society, in which behavioral
changes occur independently at each choice point, as described by the \rd for behavioral
strategies. Perceptual noise affects the realization of behavior, but not the fitness-based
selection thereof. Alternative noise-perturbed versions of the replicator dynamics in which the
fitness-based selection of strategies takes noise into account are conceivable, but we must put
the exploration of alternatives aside here. We suggest that our approach is practical,
conservative and plausible: it is practical because it is easy to implement; it is conservative
because it connects the \rdd with the \rmd (see Appendix~\ref{sec:diffusion-as-special}); it is
also plausible because the \rdd can be derived from a myopic update protocol in which agents
themselves are unaware of perceptual noise (see Appendix~\ref{sec:deriv-rdd-imit}).

@@

The \rdd is as a two-step procedure: first calculate where the replicator dynamic takes us,
then apply diffusion. To understand the \rdd better, we isolate the novel diffusion part in
Section~\ref{sec:iterated-diffusion}, before exploring how diffusion and fitness-based
selection interact in Section~\ref{sec:simulations}.

@@

Confusion of states should intuitively be a function of their
perceptual similarity: the more similar two states are, the more likely they could be mistaken for each other.
To make this concrete, let us assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. The
distance $\card{\mystate{i} - \mystate{j}}$ is the objective, physical
similarity between two states $\mystate{i}$ and
$\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \imprecision) =
      \begin{cases}
    1 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \imprecision^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\imprecision \ge 0$ is a imprecision parameter. When $\imprecision=0$ agents perfectly
discriminate between states; when $\imprecision \rightarrow \infty$ agents cannot discriminate
states at all. Figure~\ref{fig:NosofskySim} gives an impression of Nosofsky-similarity for
different parameter values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different areas of the state space,
but we stick with Nosofsky's similarity function for the time being, because it is
mathematically simple, yet an established notion in mathematical psychology.

@@

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \imprecision)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \imprecision)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\imprecision$. For $\imprecision = 0$, the confusion matrix has $C_{ii}
= 1$ and $C_{ij} = 0$ for $i \neq j$. For $\imprecision > 0$, $C$ is positive, i.e., $C_{ij} >0$ for
all $i$ and $j$. For $\imprecision \rightarrow \infty$, we find $C_{ij}
\rightarrow \nicefrac{1}{\card{\States}}$ for all $i$ and $j$.

If states are confused on average with a probability proportional to
their similarity, repeated application of the operation in
Equation~(\ref{eq:confusion-function}) causes behavioral dispositions
gradually to diffuse along a gradient of similarity of
states. Consequently, iterated diffusion leads to a smoothing out and
an eventual equalization of behavioral strategies, for both the sender
and the receiver. After $n$ steps of iterated diffusion an initial
sender strategy $\Sstrat$ will be $C^n \Sstrat$, and an initial
receiver strategy $\Rstrat$ will be $\Rstrat C^n$. If we assume that
$\imprecision > 0$, $C$ is a positive row-stochastic matrix (each state
could in principle be confused as any other state, albeit with
possibly a very low probability). It then follows from the
Perron-Frobenius theorem that the limit $C^\infty = \lim_{n
  \rightarrow \infty} C^n$ exists (it is the Perron-projection) and
all of its rows are identical. That is why, in the limit of iterated
diffusion, $C^\infty \Sstrat$ has identical rows (messages are sent
with the same probability in each state), and $\Rstrat C^\infty$
likewise has identical rows (every message is interpreted
equally). All of this is in line with the intuition that diffusion of
behavior under confusion of states iteratively equalizes behavior for
similar states; if all states can be confused for one another in
principle, behavior smoothes out entirely in the limit.

@@

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\imprecision = 0$, the
\rdd reduces to the \rd. If $\imprecision \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion process that we looked at in the previous
section. We will show presently that for reasonable in-between levels
of imprecision $\imprecision$, the \rdd can lead to reasonably
efficient, yet vague signaling behavior. Suitable levels of
imprecision can have further accelerating and, surprisingly, unifying
effects on meaning evolution.

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind3098.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\imprecision = 0$}
    \label{fig:example_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind3452.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\imprecision = 0.1$}
    \label{fig:example_stratsB}
  \end{subfigure}

  % \begin{subfigure}[]{0.45\textwidth}
  %   \includegraphics[width=\textwidth]{plots/strat_example_NS-10_tol-005_imp0_ind1001.pdf}
  %   \caption{$\ns = 10$, $\toler = 0.1$, $\imprecision = 0$}
  %   \label{fig:example_stratsC}
  % \end{subfigure}
  % \hfill
  % \begin{subfigure}[]{0.45\textwidth}
  %   \includegraphics[width=\textwidth]{plots/strat_example_NS-10_tol-005_imp005_ind1201.pdf}
  %   \caption{$\ns = 10$, $\toler = 0.1$, $\imprecision = 0.05$}
  %   \label{fig:example_stratsD}
  % \end{subfigure}
